Is partner choice related to prosociality? A cross-national investigation

### 2022-09-13 manuscript.Rmd
changes to text

abstract: |
 Why does human prosociality vary around the world? Evolutionary models and laboratory experiments suggest that possibilities for partner choice (i.e. the ability to leave unprofitable relationships and strike up new ones) should promote cooperation across human societies. Leveraging the Global Preferences Survey (n = 27,125; 27 countries) and the World Values Survey (n = 54,728; 32 countries), we test this theory by estimating the associations between relational mobility, a socioecological measure of partner choice, and a wide variety of prosocial attitudes and behaviours, including impersonal altruism, reciprocity, trust, collective action, and moral judgements of antisocial behaviour. Contrary to our pre-registered predictions, we found little evidence to suggest that partner choice is related to prosociality across countries. After controlling for posited shared causes of relational mobility and prosociality around the world --- environmental harshness, subsistence style, and geographic and linguistic proximity --- we found that only impersonal altruism and trust in people from another religion are positively related to relational mobility. We did not find positive relationships between relational mobility and reciprocity, generalised trust, collective action, or moral judgements of antisocial behaviour. These findings challenge existing evolutionary theories of human cooperation which emphasise partner choice as a key explanatory mechanism, and highlight the need to generalise theoretical models and controlled experiments to global samples.

note              : "This working paper has not yet been peer-reviewed."
keywords          : "partner choice, relational mobility, cooperation, prosociality, cross-cultural"
wordcount         : "4600 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
appendix          : "appendix.Rmd"
csl               : apa.csl
urlcolor          : blue
---

@@

# Introduction

@@

One factor that could explain global variation in prosociality is differing possibilities for partner choice across countries. Here, 'partners' are defined as individuals that people socially interact with to provide mutual benefits (e.g. friends, neighbours, colleagues, mates). Theoretical models of partner choice show that when individuals can leave interactions with uncooperative partners and actively choose new interactions with cooperative partners, cooperation can evolve and be sustained [@Aktipis2004; @Aktipis2011; @Enquist1993; @Roberts1998; @Roberts2020; @Roberts2021]. Partner choice allows for the assortative matching of cooperators, creating a market in which individuals use prosocial displays to compete for access to profitable social partnerships [@Barclay2013; @Barclay2016b]. Thus, partner choice models predict that humans should be more prosocial and cooperative if they are able to leave unprofitable partnerships and freely choose new partnerships.

@@

As well as predicting behaviour in the lab and in small-scale societies, partner choice models also predict that socioecological conditions favouring partner choice should promote prosociality in countries around the world. One recently developed socioecological variable that captures differing possibilities for partner choice is relational mobility [@Yuki2012]. Relational mobility captures "how much freedom and opportunity a society affords individuals to choose and dispose of interpersonal relationships based on personal preference" [p. 7521; @Thomson2018]. In societies with low relational mobility, people do not actively choose their relationships and their social partners are relatively fixed. By contrast, in societies with high relational mobility, people actively choose who they interact with, creating dynamic social networks. The former is akin to classic partner control models in evolutionary game theory, where individuals are forced to interact for a fixed period [@Axelrod1981], whereas the latter is akin to models of partner choice and biological markets [@Barclay2013].

We hypothesise, then, that people in higher relational mobility societies should express more prosocial behaviour and attitudes. Previous work has begun to test this hypothesis. For example, research has shown that people in higher relational mobility societies provide social support to others more frequently [@Kito2017], have greater trust in strangers [@Thomson2018], and are more likely to give gifts in romantic relationships [@Komiya2019]. Conversely, a recent meta-analysis found that people in higher relational mobility societies did not contribute more in incentivised social dilemma experiments [@Spadaro_preprint]. However, social support and cooperation in social dilemmas are only a subset of the possible measures of prosocial behaviours and attitudes, which also include impersonal altruism, reciprocity, generalised trust, collective action, and moral assessments of cheating behaviour. The relationship between relational mobility and these prosocial behaviours and attitudes is less understood.

Here, we report the results of two pre-registered studies of the cross-national associations between relational mobility, our socioecological proxy for partner choice, and a range of prosocial behaviours and attitudes. In Study 1, we leveraged data from the Global Preferences Survey [@Falk2018], a cross-national study of social preferences including impersonal altruism, positive reciprocity, and generalised trust. We linked these data to country-level relational mobility scores from 27 countries [@Thomson2018]. In Study 2, we used variables from the World Values Survey [@WVS2014] measuring collective action, trust, and moral assessments of cheating behaviour, and linked these data to relational mobility scores from 32 countries [@Thomson2018]. Based on existing theory and literature, we pre-registered for both studies that we would find a positive relationship between relational mobility and prosocial behaviours and attitudes (https://osf.io/e528t/).

@@

Participants in the Global Preferences Survey were asked a series of self-report questions that measure the following social preferences: altruism, generalised trust, positive reciprocity, negative reciprocity, risk-taking, and patience. For the purposes of our study, we focused on the altruism, trust, and positive reciprocity items. Negative reciprocity was not studied, as previous factor analyses have shown that punitive behaviour forms a separate latent variable distinct from cooperation [@Peysakhovich2014].

Altruism was measured by two items; a hypothetical charitable donation ("Imagine the following situation: Today you unexpectedly received 1000 euros. How much of this amount would you donate to a good cause?") and willingness to unconditionally donate to charity ("How willing are you to give to good causes without expecting anything in return?"). Trust was measured by a single item: agreement with the statement "I assume that people have only the best intentions". Positive reciprocity was measured by two items: stating the price of a hypothetical thank-you gift the participant would give to a stranger who helped them, and agreement with the statement "When someone does me a favour I am willing to return it". These items have been shown to reliably predict altruistic, trusting, and reciprocal behaviour in incentivised economic decision-making experiments [@Falk2016].

@@

We related measures of prosociality from the Global Preferences Survey to country-level relational mobility latent scores [@Thomson2018]. Country-level data on relational mobility were retrieved from a separate multi-country study [@Thomson2018], in which 16,939 participants across 39 countries were contacted via an online survey between 2014 and 2016. We leveraged these data since they provide valid and reliable indicators of relational mobility across multiple countries. Country-level relational mobility latent scores were estimated from self-report ratings of the relational mobility of participants' immediate societies, from a previously validated scale [@Yuki2007]. Measurement equivalence analyses have shown that the scale has partial scalar invariance across countries. Positive correlations with related variables, like job mobility and number of new acquaintances, also indicate that the scale has high convergent validity [@Thomson2018].

@@

Second, we controlled for geographic and linguistic proximity between countries. Countries that are close to one another and share common cultural ancestors are likely to be more similar to one another, due to similar ecologies, climates, institutions, and norms (see Figure \@ref(fig:plotDAG)). To account for these unmeasured confounds, we allowed countries to covary according to geographic and linguistic proximity in our models. Geographic proximity was calculated as the inverse of the logged geodesic distance between country capital cities [data from the R package _maps_; @Brownrigg2018] using the R package _geosphere_ [@Hijmans2019]. Linguistic proximity between two countries was calculated as the cultural proximity between all languages spoken within those countries, weighted by speaker percentages [see Supplementary Methods for more details; @Glottolog; @Ethnologue].

@@

To estimate the cross-national relationships between prosociality and relational mobility, we fitted pre-registered Bayesian multilevel regression models to the data (https://osf.io/e528t/). We analysed the data in long format, with multiple prosociality measures per participant (n = `r printnum(as.numeric(nrow(d1)), digits = 0)`). The outcome variable was the score for the particular prosociality measure. The country-level predictor variable was the relational mobility latent score, with latent standard deviations included in the model to account for measurement error. We included random intercepts for participants, and random intercepts and slopes for prosociality measures (altruism, trust, and positive reciprocity) and countries (see Supplementary Methods). In order to systematically compare the various effects of our variables and controls, we fitted several models: (1) an intercept-only model, (2) a model including relational mobility as a predictor, and (3) a model additionally controlling for environmental harshness and subsistence type. In all models, we allowed country random intercepts to covary according to geographic and linguistic proximity. We used approximate leave-one-out cross-validation to compare models [@Vehtari2017].

@@

(ref:plotGPS1Caption) _Posterior predictions from a Bayesian multilevel regression predicting prosocial preferences from country-level relational mobility, without control variables._ (a) The overall effect of relational mobility on prosociality. (b-d) The individual effects of relational mobility on altruism, positive reciprocity, and generalised trust. Lines and shaded areas indicate median posterior regression lines and 95% credible intervals. Points indicate average prosociality levels and relational mobility scores for each of the 27 countries, with error bars representing +/- 1 standard error.

@@

We also included two additional predictors as control variables: environmental harshness and subsistence style. Model comparison revealed that additionally conditioning on both environmental harshness and subsistence style improved model fit over a model containing only relational mobility (difference in expected log predictive density = `r printnum(abs(looCompare1b[2,1]))`, standard error = `r printnum(abs(looCompare1b[2,2]))`). The median posterior slope for relational mobility predicting overall prosocial preferences was `r printnum(median(slope1))`, 95% credible interval [`r printnum(quantile(slope1, 0.025))` `r printnum(quantile(slope1, 0.975))`] (Figure \@ref(fig:plotGPS2)). Incorporating random effects further revealed that relational mobility now slightly positively predicted altruism (median posterior slope = `r printnum(median(slope2))`, 95% CI [`r printnum(quantile(slope2, 0.025))` `r printnum(quantile(slope2, 0.975))`]), did not predict positive reciprocity (median posterior slope = `r printnum(median(slope3))`, 95% CI [`r printnum(quantile(slope3, 0.025))` `r printnum(quantile(slope3, 0.975))`]), and _negatively_ predicted generalised trust (median posterior slope = `r printnum(median(slope4))`, 95% CI [`r printnum(quantile(slope4, 0.025))` `r printnum(quantile(slope4, 0.975))`]). The slight relationship between relational mobility and impersonal altruism is in line with our pre-registered hypothesis, but the negative relationship between relational mobility and generalised trust contradicts previous research suggesting that relational mobility is positively related to trust in others [@Yuki2007; @Thomson2018].

(ref:plotGPS2Caption) _Posterior predictions from a Bayesian multilevel regression predicting prosocial preferences from country-level relational mobility, controlling for environmental harshness and subsistence style._ (a) The overall effect of relational mobility on prosociality. (b-d) The individual effects of relational mobility on altruism, positive reciprocity, and generalised trust. Lines and shaded areas indicate median posterior regression lines and 95% credible intervals. Points indicate average prosociality levels and relational mobility scores for each of the 27 countries, with error bars representing +/- 1 standard error.

@@

There are several possible explanations for these mixed results. First, over half of our sample of countries were from Western Europe and North America, where relational mobility is higher than average. This does not leave much variation to detect associations, especially with a small sample size of 27 countries. Second, only a small set of prosociality measures were available in the Global Preferences Survey, limited to charitable donations, exchanges of gifts and favours, and generalised trust. As such, this dataset did not cover other important aspects of prosociality, such as collective action and moral judgements of antisocial cheating behaviour.

In order to investigate whether these factors could explain our results, we conducted a second study with a different dataset. In Study 2, we leveraged data from the World Values Survey [@WVS2014], a multi-country self-report study of values and attitudes. This study has global coverage and includes items measuring a wide variety of prosocial behaviours and attitudes. We were able to link data from 32 countries to country-level data on relational mobility, expanding our sample size and including additional Asian countries. We hypothesised that individuals from countries with higher relational mobility would: (1) be more likely to belong to humanitarian and charitable organisations, our measure of collective action; (2) be more likely to believe that most people can be trusted; (3) report higher trust for specific groups; and (4) report lower justifiability for self-interested moral transgressions.

@@

Participants in both the World Values Survey and the European Values Survey answer a range of self-report questions on social values, societal wellbeing, trust, economic values, religion, politics, and ethics. For the purposes of our study, we highlighted several variables as measures of cooperation, trust, and prosociality. The first variable captures cooperation via collective action ["Are you a member of a charitable or humanitarian organisation?"; for a similar interpretation of this variable, see @Jacquet2021]. The second variable captures generalised trust ("Generally speaking, would you say that most people can be trusted or that you need to be very careful in dealing with people?"). The third set of variables captures levels of trust in specific groups of people, namely family, neighborhood, personal acquaintances, people the respondent has met for the first time, people of another religion, and people of another nationality. The fourth set of variables captures the justifiability of different self-interested moral trangressions, including claiming unentitled government benefits, avoiding a fare on public transport, cheating on taxes, and someone accepting a bribe.

@@

As described in Study 1, we included measurement error on the relational mobility latent scores and accounted for spatial and cultural non-independence between countries with correlated random intercepts. We additionally controlled for environmental harshness and subsistence style. All analyses were conducted in R v4.0.2. [@R].

@@

For our measure of cooperation and collective action --- charitable organisation membership --- model comparison revealed that adding relational mobility as a predictor improved model fit over a null intercept-only model (difference in expected log predictive density = `r printnum(abs(looCompare5[2,1]))`, standard error = `r printnum(abs(looCompare5[2,2]))`). The posterior log odds slope for relational mobility predicting charitable organisation membership was in the expected direction, but the 95% credible interval included zero (median posterior slope = `r printnum(median(slope1))`, 95% CI [`r printnum(quantile(slope1, 0.025))` `r printnum(quantile(slope1, 0.975))`]; Figure \@ref(fig:plotWVS5a)). The 95% credible interval continued to include zero after controlling for environmental harshness and subsistence type (median posterior slope = `r printnum(median(slope2))`, 95% CI [`r printnum(quantile(slope2, 0.025))` `r printnum(quantile(slope2, 0.975))`]; Supplementary Figure \@ref(fig:plotWVS5b)).

(ref:plotWVS5aCaption) _Posterior predictions from a Bayesian multilevel logistic regression predicting charitable organisation membership from country-level relational mobility, without controls._ The line and shaded area indicate the median posterior regression line and 95% credible intervals. Points indicate the proportion of individuals belonging to charitable organisations on the y-axis and relational mobility scores on the x-axis, for each of the 32 countries, with error bars representing +/- 1 standard error.

@@

For generalised trust, model comparison revealed that adding relational mobility as a predictor improved model fit over a null intercept-only model (difference in expected log predictive density = `r printnum(abs(looCompare6[2,1]))`, standard error = `r printnum(abs(looCompare5[2,2]))`). The 95% credible interval for the posterior log odds slope for relational mobility predicting generalised trust included zero (median posterior slope = `r printnum(median(slope1))`, 95% CI [`r printnum(quantile(slope1, 0.025))` `r printnum(quantile(slope1, 0.975))`]; Figure \@ref(fig:plotWVS6a)). The 95% credible interval continued to include zero after controlling for environmental harshness and subsistence type (median posterior slope = `r printnum(median(slope2))`, 95% CI [`r printnum(quantile(slope2, 0.025))` `r printnum(quantile(slope2, 0.975))`]; Supplementary Figure \@ref(fig:plotWVS6b)).

(ref:plotWVS6aCaption) _Posterior predictions from a Bayesian multilevel logistic regression predicting generalised trust from country-level relational mobility, without controls._ The line and shaded area indicate the median posterior regression line and 95% credible intervals. Points indicate the proportion of individuals stating that "most people can be trusted" on the y-axis and relational mobility scores on the x-axis, for each of the 32 countries, with error bars representing +/- 1 standard error.

>>>>

abstract: |
 Why does human prosociality vary around the world? Evolutionary models and laboratory experiments suggest that possibilities for partner choice (i.e., the ability to leave unprofitable relationships and strike up new ones) should promote cooperation across human societies. Leveraging the Global Preferences Survey (n = 27,125; 27 countries) and the World Values Survey (n = 54,728; 32 countries), we test this theory by estimating the associations between relational mobility, a socioecological measure of partner choice, and a wide variety of prosocial attitudes and behaviours, including impersonal altruism, reciprocity, trust, collective action, and moral judgements of antisocial behaviour. Contrary to our pre-registered predictions, we found little evidence to suggest that partner choice is related to prosociality across countries. After controlling for posited shared causes of relational mobility and prosociality around the world --- environmental harshness, subsistence style, and geographic and linguistic proximity --- we found that only impersonal altruism and trust in people from another religion are positively related to relational mobility. We did not find positive relationships between relational mobility and reciprocity, generalised trust, collective action, or moral judgements of antisocial behaviour. These findings challenge existing evolutionary theories of human cooperation which emphasise partner choice as a key explanatory mechanism, and highlight the need to generalise theoretical models and controlled experiments to global samples.

note              : "This working paper has not yet been peer-reviewed."
keywords          : "partner choice, relational mobility, cooperation, prosociality, cross-cultural"
wordcount         : "5500 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
appendix          : "appendix.Rmd"
csl               : apa.csl
urlcolor          : blue
---

@@

@@

One factor that could explain global variation in prosociality is differing possibilities for partner choice across countries. Here, 'partners' are defined as individuals that people socially interact with to provide mutual benefits (e.g., friends, neighbours, colleagues, mates). Theoretical models of partner choice show that when individuals can leave interactions with uncooperative partners and actively choose new interactions with cooperative partners, cooperation can evolve and be sustained [@Aktipis2004; @Aktipis2011; @Enquist1993; @Roberts1998; @Roberts2020; @Roberts2021]. Partner choice allows for the assortative matching of cooperators, creating a market in which individuals use prosocial displays to compete for access to profitable social partnerships [@Barclay2013; @Barclay2016b]. Thus, partner choice models predict that humans should be more prosocial and cooperative if they are able to leave unprofitable partnerships and freely choose new partnerships.

@@

As well as predicting behaviour in the lab and in small-scale societies, partner choice models also predict that socioecological conditions favouring partner choice should promote prosociality in countries around the world. One recently developed socioecological variable that captures differing possibilities for partner choice is relational mobility [@Yuki2012]. Relational mobility captures "how much freedom and opportunity a society affords individuals to choose and dispose of interpersonal relationships based on personal preference" [p. 7521, @Thomson2018]. In societies with low relational mobility, people do not actively choose their relationships and their social partners are relatively fixed. By contrast, in societies with high relational mobility, people actively choose who they interact with, creating dynamic social networks.

Societies with low levels of relational mobility are akin to classic partner control models in evolutionary game theory, where individuals are forced to interact for a fixed period [@Axelrod1981]. Partner control _can_ promote prosocial behaviour, but only on the condition that individuals are able to reward their partners' cooperative acts and effectively punish defection. By contrast, societies with high levels of relational mobility are akin to models of partner choice and biological markets [@Barclay2013], which promote the evolution of cooperation under a potentially wider range of conditions than partner control models [@Aktipis2004; @Aktipis2011; @Enquist1993; @Roberts1998; @Roberts2020; @Roberts2021]. Indeed, @Barclay2016a found that people behave more prosocially when they can leave uncooperative partners compared to when they are forced to interact with them over fixed periods, even with the possibility of reciprocation and punishment.

We hypothesise, then, that people in higher relational mobility societies should express more prosocial behaviour and attitudes. Previous work has begun to test this hypothesis. For example, research has shown that people in higher relational mobility societies provide social support to others more frequently [@Kito2017], have greater trust in strangers [@Thomson2018], and are more likely to give gifts in romantic relationships [@Komiya2019]. Conversely, a recent meta-analysis found that people in higher relational mobility societies did not contribute more in incentivised social dilemma experiments [@Spadaro_preprint]. However, this previous work has focused on only a subset of possible measures of prosocial behaviours and attitudes: social support and cooperation in social dilemmas. Other kinds of prosociality predicted to increase under high levels of relational mobility include impersonal altruism, reciprocity, generalised trust, collective action, and moral assessments of cheating behaviour. In addition, previous research has not studied the nature of the relationship between relational mobility and prosociality. While theoretical work has generally shown that partner choice promotes the evolution of cooperation, in some models too much partner choice is actually harmful for cooperation, because partner choice reduces interdependence with one's current partner [@Barclay2020] and defectors can easily find new individuals to exploit [@Aktipis2004]. It is thus possible that the positive relationship between relational mobility and prosociality could have a "hump-backed" shape, whereby relational mobility initially increases prosociality but too much relational mobility decreases it.

Here, we report the results of two pre-registered studies of the cross-national associations between relational mobility, our socioecological proxy for partner choice, and a range of prosocial behaviours and attitudes. In Study 1, we leveraged data from the Global Preferences Survey [@Falk2018], a cross-national study of social preferences including impersonal altruism, positive reciprocity, and generalised trust. We focused on these particular measures of prosociality because altruistic, reciprocal, and trusting behaviours have been shown to reflect a single behavioural construct dubbed the "cooperative phenotype" in previous work [@Peysakhovich2014]. All three of these behaviours are predicted to increase under higher levels of partner choice: altruistic and reciprocal prosocial behaviours become useful as signals of cooperative intent for potential partners, and levels of trust thus increase along with levels of prosociality in the population. In Study 2, we used variables from the World Values Survey [@WVS2014] measuring collective action, moral assessments of cheating behaviour, and trust, which additionally capture people's prosocial contribution to social dilemmas and willingness to uphold prosocial moral norms.

Across both studies, we linked these prosociality data to relational mobility scores from a previous international survey [@Thomson2018]. Based on existing theory and literature, we pre-registered for both studies that we would find positive linear relationships between relational mobility and prosocial behaviours and attitudes: as relational mobility increases around the world, so should prosociality (https://osf.io/e528t/). In addition to our pre-registered analyses, we also explored potential non-linear relationships between relational mobility and prosocial behaviour and attitudes.

@@

Participants in the Global Preferences Survey were asked a series of self-report questions that measure the following social preferences: altruism, generalised trust, positive reciprocity, negative reciprocity, risk-taking, and patience. For the purposes of our study, we focused on the altruism, trust, and positive reciprocity items (for raw country-level data, see Supplementary Table \@ref(tab:tableRaw1)). Negative reciprocity was not studied, as previous factor analyses have shown that punitive behaviour forms a separate latent variable distinct from cooperation [@Peysakhovich2014].

Altruism was measured by two items; a hypothetical charitable donation ("Imagine the following situation: Today you unexpectedly received 1000 euros. How much of this amount would you donate to a good cause?") and willingness to unconditionally donate to charity ("How willing are you to give to good causes without expecting anything in return?"). Trust was measured by a single item: agreement with the statement "I assume that people have only the best intentions". Positive reciprocity was measured by two items: stating the price of a hypothetical thank-you gift the participant would give to a stranger who helped them, and agreement with the statement "When someone does me a favour I am willing to return it". These items have been shown to reliably predict altruistic, trusting, and reciprocal behaviour in incentivised economic decision-making experiments [@Falk2016]. These items also have metric invariance across countries, suggesting that they can be meaningfully compared around the world (Supplementary Table \@ref(tab:tableInvarianceGPS)).

@@

We related measures of prosociality from the Global Preferences Survey to country-level relational mobility latent scores [@Thomson2018]. Country-level data on relational mobility were retrieved from a separate multi-country study [@Thomson2018], in which 16,939 participants across 39 countries were contacted via an online survey between 2014 and 2016. We leveraged these data since they provide valid and reliable indicators of relational mobility across multiple countries. Country-level relational mobility latent scores were estimated from self-report ratings of the relational mobility of participants' immediate societies, from a previously validated scale [@Yuki2007]. Measurement invariance analyses have shown that the scale has partial scalar invariance across countries. Positive correlations with related variables, like job mobility and number of new acquaintances, also indicate that the scale has high convergent validity [@Thomson2018].

@@

Second, we controlled for geographic and linguistic proximity between countries. Countries that are close to one another and share common cultural ancestors are likely to be more similar to one another, due to similar ecologies, climates, institutions, and norms (see Figure \@ref(fig:plotDAG)). To account for these unmeasured confounds, we allowed countries to covary according to geographic and linguistic proximity in our models. Geographic proximity was calculated as the inverse of the logged geodesic distance between country capital cities [data from the R package _maps_, @Brownrigg2018] using the R package _geosphere_ [@Hijmans2019]. Linguistic proximity between two countries was calculated as the cultural proximity between all languages spoken within those countries, weighted by speaker percentages [@Glottolog; @Ethnologue]: see Supplementary Methods for more details. While we acknowledge that linguistic proximity might not capture all cultural similarities between nations (e.g., differences in social structure and institutions between majority Spanish-speaking countries), this proxy for cultural relatedness has been shown to predict a number of national-level outcomes around the world, such as democratic institutions [@Matthews2016] and human development [@Claessens_preprint].

@@

```{r echo=FALSE}
tar_load(power1)
```

To estimate the cross-national relationships between prosociality and relational mobility, we fitted pre-registered Bayesian multilevel regression models to the data (https://osf.io/e528t/). We analysed the data in long format, with multiple prosociality measures per participant (n = `r printnum(as.numeric(nrow(d1)), digits = 0)`). The outcome variable was the score for the particular prosociality measure. The country-level predictor variable was the relational mobility latent score, with latent standard deviations included in the model to account for measurement error. We included random intercepts for participants and countries, and random intercepts and slopes for prosociality measures (altruism, trust, and positive reciprocity; see Supplementary Methods). This multilevel structure deals with the fact that some countries have more observations than others, weighting the population-level estimates accordingly.

In order to systematically compare the various effects of our variables and controls, we fitted several models: (1) an intercept-only model, (2) a model including relational mobility as a predictor, (3) a model additionally controlling for environmental harshness and subsistence type, and (4) a model including controls and a quadratic effect of relational mobility. In all models, we allowed country random intercepts to covary according to geographic and linguistic proximity. Power analysis simulations revealed that the model with controls would be able to detect a medium effect of relational mobility ($\beta$ = `r str_sub(power1$description[2], -4)`) with `r printnum((power1$x / power1$n)*100, digits = 0)`% power (Supplementary Table \@ref(tab:tablePower)). We used approximate leave-one-out cross-validation to compare models [@Vehtari2017].

@@

(ref:plotGPS1Caption) _Posterior predictions from a Bayesian multilevel regression predicting prosocial preferences from country-level relational mobility, without control variables._ (a) The overall effect of relational mobility on prosociality. (b-d) The individual effects of relational mobility on altruism, positive reciprocity, and generalised trust. Lines and shaded areas indicate median posterior regression lines and 95% credible intervals. Points indicate average prosociality levels and relational mobility scores for each of the 27 countries, with error bars representing +/- 1 standard error. Letters represent country ISO codes.

@@

We also included two additional predictors as control variables: environmental harshness and subsistence style. Model comparison revealed that additionally conditioning on both environmental harshness and subsistence style improved model fit over a model containing only relational mobility (difference in expected log predictive density = `r printnum(abs(looCompare1b[2,1]))`, standard error = `r printnum(abs(looCompare1b[2,2]))`). The median posterior slope for relational mobility predicting overall prosocial preferences was `r printnum(median(slope1))`, 95% credible interval [`r printnum(quantile(slope1, 0.025))` `r printnum(quantile(slope1, 0.975))`] (Figure \@ref(fig:plotGPS2)). Incorporating random effects further revealed that relational mobility now slightly positively predicted altruism (median posterior slope = `r printnum(median(slope2))`, 95% CI [`r printnum(quantile(slope2, 0.025))` `r printnum(quantile(slope2, 0.975))`]), did not predict positive reciprocity (median posterior slope = `r printnum(median(slope3))`, 95% CI [`r printnum(quantile(slope3, 0.025))` `r printnum(quantile(slope3, 0.975))`]), and _negatively_ predicted generalised trust (median posterior slope = `r printnum(median(slope4))`, 95% CI [`r printnum(quantile(slope4, 0.025))` `r printnum(quantile(slope4, 0.975))`]). The slight relationship between relational mobility and impersonal altruism is in line with our pre-registered hypothesis, but the negative relationship between relational mobility and generalised trust contradicts previous research suggesting that relational mobility is positively related to trust in others [@Yuki2007; @Thomson2018]. There was no quadratic effect of relational mobility in the model including controls (Supplementary Table \@ref(tab:tableQuadratic)).

(ref:plotGPS2Caption) _Posterior predictions from a Bayesian multilevel regression predicting prosocial preferences from country-level relational mobility, controlling for environmental harshness and subsistence style._ (a) The overall effect of relational mobility on prosociality. (b-d) The individual effects of relational mobility on altruism, positive reciprocity, and generalised trust. Lines and shaded areas indicate median posterior regression lines and 95% credible intervals. Points indicate average prosociality levels and relational mobility scores for each of the 27 countries, with error bars representing +/- 1 standard error. Letters represent country ISO codes.

@@

There are several possible explanations for these mixed results. First, over half of our sample of countries were from Western Europe and North America, where relational mobility is higher than average. This does not leave much variation to detect associations, especially with a small sample size of 27 countries. Second, only a small set of prosociality measures were available in the Global Preferences Survey, limited to charitable donations, exchanges of gifts and favours, and generalised trust. As such, this dataset did not cover other important aspects of prosociality, such as prosocial contributions to social dilemmas and willingness to uphold prosocial norms.

In order to investigate whether these factors could explain our results, we conducted a second study with a different dataset. In Study 2, we leveraged data from the World Values Survey [@WVS2014], a multi-country self-report study of values and attitudes. This study has global coverage and includes items measuring a wide variety of prosocial behaviours and attitudes. We were able to link data from 32 countries to country-level data on relational mobility, expanding our sample size and including additional Asian countries. We hypothesised that individuals from countries with higher relational mobility would be more likely to belong to humanitarian and charitable organisations, our measure of collective action and prosocial contribution to social dilemmas, and more likely to report that violations of prosocial norms are morally unjustifiable. Both of these are indirect measures of cooperative and prosocial behaviours that could feasibly provide signals of cooperative intent in biological markets. Repeating the prediction from our first study, we also hypothesise that individuals from countries with higher relational mobility will show higher levels of trust in others.

@@

Participants in both the World Values Survey and the European Values Survey answer a range of self-report questions on social values, societal wellbeing, trust, economic values, religion, politics, and ethics. For the purposes of our study, we highlighted several variables as measures of cooperation, trust, and prosociality. The first variable captures cooperation via collective action: "Are you a member of a charitable or humanitarian organisation?" For a similar interpretation of this variable, see @Jacquet2021. The second variable captures generalised trust: "Generally speaking, would you say that most people can be trusted or that you need to be very careful in dealing with people?". The third set of variables captures levels of trust in specific groups of people, namely family, neighborhood, personal acquaintances, people the respondent has met for the first time, people of another religion, and people of another nationality. The fourth set of variables captures the justifiability of different self-interested moral trangressions, including claiming unentitled government benefits, avoiding a fare on public transport, cheating on taxes, and someone accepting a bribe. Both the set of items measuring trust in different groups and the set of items measuring moral justifiability for different moral transgressions have metric invariance across countries, suggesting that they can be meaningfully compared around the world (Supplementary Tables \@ref(tab:tableInvarianceWVS1) and \@ref(tab:tableInvarianceWVS2)).

@@

As described in Study 1, we included measurement error on the relational mobility latent scores and accounted for spatial and cultural non-independence between countries with correlated random intercepts. We additionally fitted models that controlled for environmental harshness and subsistence style and included a quadratic effect of relational mobility. Power analysis simulations revealed that the models with controls would be able to detect small-to-medium effects of relational mobility with roughly 80% power (Supplementary Table \@ref(tab:tablePower)). All analyses were conducted in R v4.0.2. [@R].

@@

For our measure of cooperation and collective action --- charitable organisation membership --- model comparison revealed that adding relational mobility as a predictor improved model fit over a null intercept-only model (difference in expected log predictive density = `r printnum(abs(looCompare5[2,1]))`, standard error = `r printnum(abs(looCompare5[2,2]))`). The posterior log odds slope for relational mobility predicting charitable organisation membership was in the expected direction, but the 95% credible interval included zero (median posterior slope = `r printnum(median(slope1))`, 95% CI [`r printnum(quantile(slope1, 0.025))` `r printnum(quantile(slope1, 0.975))`]; Figure \@ref(fig:plotWVS5a)). The 95% credible interval continued to include zero after controlling for environmental harshness and subsistence type (median posterior slope = `r printnum(median(slope2))`, 95% CI [`r printnum(quantile(slope2, 0.025))` `r printnum(quantile(slope2, 0.975))`]; Supplementary Figure \@ref(fig:plotWVS5b)). There was no quadratic effect of relational mobility on charitable organisation membership (Supplementary Table \@ref(tab:tableQuadratic)).

(ref:plotWVS5aCaption) _Posterior predictions from a Bayesian multilevel logistic regression predicting charitable organisation membership from country-level relational mobility, without controls._ The line and shaded area indicate the median posterior regression line and 95% credible intervals. Points indicate the proportion of individuals belonging to charitable organisations on the y-axis and relational mobility scores on the x-axis, for each of the 32 countries, with error bars representing +/- 1 standard error. Letters represent country ISO codes.

@@

For generalised trust, model comparison revealed that adding relational mobility as a predictor improved model fit over a null intercept-only model (difference in expected log predictive density = `r printnum(abs(looCompare6[2,1]))`, standard error = `r printnum(abs(looCompare5[2,2]))`). The 95% credible interval for the posterior log odds slope for relational mobility predicting generalised trust included zero (median posterior slope = `r printnum(median(slope1))`, 95% CI [`r printnum(quantile(slope1, 0.025))` `r printnum(quantile(slope1, 0.975))`]; Figure \@ref(fig:plotWVS6a)). The 95% credible interval continued to include zero after controlling for environmental harshness and subsistence type (median posterior slope = `r printnum(median(slope2))`, 95% CI [`r printnum(quantile(slope2, 0.025))` `r printnum(quantile(slope2, 0.975))`]; Supplementary Figure \@ref(fig:plotWVS6b)). There was no quadratic effect of relational mobility on generalised trust (Supplementary Table \@ref(tab:tableQuadratic)).

(ref:plotWVS6aCaption) _Posterior predictions from a Bayesian multilevel logistic regression predicting generalised trust from country-level relational mobility, without controls._ The line and shaded area indicate the median posterior regression line and 95% credible intervals. Points indicate the proportion of individuals stating that "most people can be trusted" on the y-axis and relational mobility scores on the x-axis, for each of the 32 countries, with error bars representing +/- 1 standard error. Letters represent country ISO codes.
